---
title: "LaTeX and Knitr"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: true
---

<br>

# Description

My goal here is to get you up and running with LaTeX and `knitr`. Because _up and running_ is my goal, there are many topics related to LaTeX and `knitr` (especially LaTeX) that I do not cover here—for instance: tables and figures. However, I think this guide should get you pretty well along your track. If you run into issues, try the resources I've provided, try Google, and don't be afraid to ask questions. Also, let me know if you think there is something that definitely should be added to these notes.

# Installation

## LaTeX

The very first thing you need to do is [download a LaTeX distribution](https://www.latex-project.org/get/) for your operating system:

- The standard distribution for Macs (OSX) is MacTex.
- Windows users can choose between MiKTeX or TeX Live (or proTeXt). I don't think it really matters which one you choose. I used MiKTeX in the past and was quite happy.
- Linux users probably already know what they are doing.

__Note:__ There are several solid online services available, but please install a LaTeX distribution on your machine.

__Another note:__ You might be tempted by the simplicity of the LyX world. Try to resist the temptation. LyX is a nice program, but you will inevitably want more customization than it offers, and then you will regret you didn't spend some time learning LaTeX. Give yourself some credit and go big.

## `knitr`

Next, head back over to RStudio and install the `knitr` package.^[`install.packages("knitr")`]

## `knitr` in RStudio

Now we will tell RStudio to use `knitr` to build PDF documents. Open the settings in RStudio (Menu bar » Tools » Global Options^[or command + comma (⌘+,) on a Mac]). Go to the "Sweave" tab.

1. Set the option "Weave Rnw files using" to "knitr".
2. Uncheck the box next to "Always enable Rnw concordance...".

# Getting started with LaTeX

## Pronunciation

First things first: how do you pronounce _LaTeX_? According to the [LaTeX project website](https://www.latex-project.org/about/) and to the founders of [LaTeX](https://tex.stackexchange.com/questions/17502/what-is-the-correct-pronunciation-of-tex-and-latex), there are two "preferred" pronunciations: «lah-tech» or «lay-tech».

## Resources

I'm going to walk you through a few LaTeX examples, but there is __a lot__ more out there. Besides your normal resources (Google + StackExchange/Overflow), here are a few handy resources:

- [ShareLaTeX](https://www.sharelatex.com/learn/) has a great resources section. A great place to start learning LaTeX or refresh on LaTeX specifics.
- [A LaTeX wiki book](https://en.wikibooks.org/wiki/LaTeX) with some helpful sections. Not always the easiest place to find what you want, but still a nice compilation of information.
- [The LaTeX Project](https://www.latex-project.org/help/links/) has links to other resources. The LaTeX Project also has a [brief introduction to LaTeX](https://www.latex-project.org/about/).

## Overview

LaTeX is essentially two things:

1. a mark-up language that allows you to specify the formatting you desire;
2. an engine that typesets a .tex document and creates a (beautiful) .pdf.

It is particularly good for creating large documents with tables and figures—three things with which other word processors often struggle.

To create a PDF from your .tex document, you need to open the .tex file in a program that can compile. When you installed LaTeX, it probably installed a program to edit and compile (_e.g._, Gummi, TeXShop, Texmaker, TeXstudio, or TeXworks).^[There are a lot of options. I really like [Atom](https://atom.io), but it requires more setting up. Three guides for setting up Atom with LaTeX:  [1](https://rolflekang.com/writing-latex-in-atom/), [2](http://economistry.com/2016/02/create-first-pdf-latex-atom-mac/), [3]((http://blog.matteomerola.me/2016/02/atom-for-latex)).] Or... use RStudio!

## Bare bones example

Assuming you have properly installed LaTeX and an editor/compiler (again, you can use RStudio), you can now create your first LaTeX document.

A LaTeX document needs _at least_ the following three lines of text:
```asis
\documentclass{article}
\begin{document}
\end{document}
```
However, these three lines of text will produce an empty document, so don't try to compile just yet.

The first line (`\documentclass{article}`) tells LaTeX which type of document you want to create (similar to selecting templates in Word). For class, papers, and publications, you will generally use `article`. Other options include `report`, `book`, `letter`, and `beamer`. The `beamer` option creates the slides you often see in seminars. There are many other options you can change by including them in square brackets before the curly brackets, _i.e._
```asis
\documentclass[option1, option2, ...]{article}
```
Among these options are the language and the default font size (the default is 10pt).
```asis
\documentclass[english, 12pt]{article}
```

Everything between the `\documentclass` declaration and the beginning of the document (`\begin{document}`) is known as the _preamble_—it is where you load packages, change options/settings, _etc._ You write your text (paragraphs, equations, figures, tables) in between the calls of `\begin{document}` and `\end{document}` (in the document).

Let's use a package called `inputenc`. We need to load the package, similar to the way we load a package in R. It's pretty standard to use `\usepackage[utf8]{inputenc}` for the document's encoding, so we will insert this line of code into our preamble.
```asis
\documentclass{article}
\usepackage[utf8]{inputenc}
\begin{document}
\end{document}
```

Finally, we should probably have some text in our document.
```asis
\documentclass{article}
\usepackage[utf8]{inputenc}
\begin{document}
What a beautiful PDF.
\end{document}
```

Now, let's finally compile the document.

1. Open a new R script in RStudio (ctrl/cmd+shift+N).
2. Paste the 5 lines of code from the block above into your new R script.
3. Save the script (ctrl/cmd+s) with these five lines of code as a .tex document (_e.g._ example.tex). RStudio will probably ask if you really want to use ".tex" rather than ".R". You do: say yes.
4. In the toolbar above your .tex file, you should see the words "Compile PDF". You can also use the menu via "File » Compile PDF" or the keyboard shortcut ctrl/cmd+shift+K.

You should now have a stunning PDF.

# LaTeX basics

You will probably want a little more out of LaTeX than we have done so far.

## Title, author, date

You add the title, author(s), and date to a LaTeX document inside of the preamble (think of them as settings for the whole document) and then tell LaTeX to print them inside the document with the `\maketitle` command (see below). LaTeX provides functions for each of these items, so we can insert them straight into our preamble:
```asis
\documentclass{article}

\usepackage[utf8]{inputenc}

\title{My beautiful PDF document}
\author{Ed Rubin}
\date{February 2017}

\begin{document}

\maketitle

What a beautiful PDF.

\end{document}
```
Update your .tex document and compile again. What happens when you leave `date{}` empty?^[Answer: There will be no date printed.] What if you leave `\date{}` off completely?^[Answer: LaTeX prints today's date.]

Let's pretend our beautiful document is supposed to have a title page. Easy:
```asis
\documentclass{article}

\usepackage[utf8]{inputenc}

\title{My beautiful PDF document}
\author{Ed Rubin}
\date{February 2017}

\begin{document}

\begin{titlepage}
\maketitle
\end{titlepage}

What a beautiful PDF.

\end{document}
```
As you are likely beginning to see, there are many commands in the LaTeX _language_ that make use of this `\begin{command}` and `\end{command}` convention.

## Text

While writing, within the document you have access to a bunch of options for text.

- `\textbf{blah}` will print the word "blah" as __bold__ text.
- `\textit{bleh}` will print the word "bleh" as _italic_ text.^[Some people use the command `\emph{}` to get italics, which is not a great idea. Depending on the packages and document class, `\emph{}` can produce different styles (italic, bold, _etc._).]
- `\underline{blih}` will print the word "blih" underlined.
- For quotations, you need to use double backticks (```  `` ```) at the start and double quotations marks at the end of the quote (`''`). Example: ``` ``This is a quote.'' ``` It is a little annoying—especially when you copy and paste from another source—but if you don't do it, your qutoation marks will be backwards half of the time.
- Skipping a line between lines of text will start a new paragraph. The command `\par` will also start a new paragraph.
- The characters `\\` will start a new line but __not__ a new paragraph (no indentation).
- If you start a new line of text (in your editor) without `\\` or without `\par`, LaTeX will ignore the new line and keep the text going where the last line ended.
- You can tell an individual paragraph not to indent by using the `\noindent` command at the beginning of the paragraph. You can tell LaTeX to never indent paragraphs by placing the command `\setlength\parindent{0pt}` into the preamble.
- To change the line spacing (the default is single spacing), you can use the `setspace` package and its functions `\singlespacing`, `\one­half­s­pac­ing`, or `\doublespacing`, _e.g._ add the following lines to your preamble:
```asis
\usepackage{setspace}
\onehalfspacing
```

## Headings

To create headings, use the built-it commands:

- `\section{}`, `\subsection{}`, `\subsubsection{}`, `\paragraph{}`, `\subparagraph{}`
- Put text in the curly brackets so your headings have text.
- Add an asterisk (`*`) before the first curly bracket to preven LaTeX from numbering the section, _e.g._ `\section{Introduction}` will have a number, while `section*{Introduction}` will not have a number.

## Lists

LaTeX provides a pretty straightforward way to make lists—numbered or unnumbered. For numbered lists, the command is `enumerate`. You combine it with `\begin{enumerate}` and `\end{enumerate}`, and in between, you use `\item` to create the individual text entries in your list.
```asis
\begin{enumerate}
  \item Raspberry
  \item Cherry
  \item Kiwi
  \item Apple
\end{enumerate}
```

Unnumbered lists work in exactly the same way, except they use the command `itemize` rather than `enumerate`.
```asis
\begin{itemize}
  \item Raspberry
  \item Cherry
  \item Kiwi
  \item Apple
\end{itemize}
```

Let's put everything we have learned so far together in our .tex document.

```asis
\documentclass[english, 11pt]{article}

\usepackage[utf8]{inputenc}

\usepackage{setspace}
\onehalfspacing

\title{My beautiful PDF document}
\author{Ed Rubin}
\date{February 2017}

\begin{document}

\begin{titlepage}
\maketitle
\end{titlepage}

\noindent What a beautiful PDF!

\section{Paragraphs, text, and lists}

\subsection{Paragraphs}

\textit{An excerpt from \underline{The War of the Worlds} (Book I), by H. G. Wells}

For the main road was a boiling stream of people, a torrent of human beings rushing
northward, one pressing on another. A great bank of dust, white and luminous in the
blaze of the sun, made everything within twenty feet of the ground grey and
indistinct and was perpetually renewed by the hurrying feet of a dense crowd of
horses and of men and women on foot, and by the wheels of vehicles of every description.

``Way!'' my brother heard voices crying. ``Make way!''

It was like riding into the smoke of a fire to approach the meeting point of the lane
and road; the crowd roared like a fire, and the dust was hot and pungent. And, indeed,
a little way up the road a villa was burning and sending rolling masses of black smoke
across the road to add to the confusion.

\subsection{Lists}

Let's make a list!

\subsubsection*{An unnumbered list of random things}

\begin{itemize}
  \item Econometrics
  \item Pastrami
  \item Garden gnomes
  \item Redwoods
\end{itemize}

\subsubsection*{A numbered list}

\noindent The hierarchy of fruit deliciousness

\begin{itemize}
  \item Raspberry
  \item Cherry
  \item Kiwi
  \item \ldots
  \item Apple
\end{itemize}

\textbf{Note:} This hierarchy does not pertain to pies.

\end{document}
```

## Math

LaTeX does math really well. However, there is some amount of learning required up front. For instance, if you want to see $\widehat{\beta} = \left(\mathbf{X}'\mathbf{X}\right)^{-1}\mathbf{X}'\mathbf{y}$, then you need to type
```asis
$\widehat{\beta} = \left(\mathbf{X}'\mathbf{X}\right)^{-1}\mathbf{X}'\mathbf{y}$
```

[ShareLaTeX has great resources](https://www.sharelatex.com/learn/Mathematical_expressions) on mathematical expressions in LaTeX, so I am going to cover the basics here and leave the rest to them.

There are two main ways (modes) in which you will want use mathematical expressions in LaTeX: inline (in sentence or paragraph) and display mode (the math gets its own line). The two modes use the same code/language to generate the mathematical expressions; the main difference is how you introduce expression to LaTeX in your document.

For instance, if you write `$y = 2^\alpha + \beta \mathbf{x}$`, LaTeX will display the equation __inline__. It will then show up in a sentence like:

My favorite formula today is $y = 2^\alpha + \beta \mathbf{x}$.

If we change the dollar signs to brackets with backslashes (_e.g._ change `$` and `$` to `\[` and `\]`), yielding `\[y = 2^\alpha + \beta \mathbf{x}\]`, then LaTeX uses display mode:

My favorite formula today is \[y = 2^\alpha + \beta \mathbf{x}\]

# `knitr`

Finally. Let's talk about `knitr`.

## Description

As you have already seen, `knitr` is a package in R. Specifically, `knitr` provides a way to create dynamic and reproducible documents that combine the typesetting aesthetics of LaTeX with the computing resources of R. LaTeX is great at producing documents but not great for statistical computations. R, without `knitr`, is great at statistics but struggles to create a report. `knitr` gives us a link between the two programs that builds beautiful PDFs whose graphs, tables, and general results dynamically update using R.

Because `knitr` builds upon LaTeX, everything you know and use in LaTeX carries over to `knitr`—`knitr` just adds a few new items.

## First steps

If you were able to compile the .tex files above, you should be set up for `knitr`.

Let's start a new document in RStudio (new R script—ctrl/cmd+shift+N).

Add the standard LaTeX preamble, document setup, and some filler text:
```asis
\documentclass[english, 11pt]{article}
\usepackage[utf8]{inputenc}

\begin{document}

Clever text.

\end{document}
```
Now save the document. Instead of saving this file as a ".tex", we are now going to save it as ".Rnw" (_e.g._ example.Rnw).^[In case you are wondering, the suffix ".Rnw" [apparently](http://kbroman.org/knitr_knutshell/pages/latex.html) stands for "R noweb"] Your average LaTeX editor/compiler will not be able to work with .Rnw files, but RStudio loves them^[To the extent that a computer program can love anything].

## LaTeX meets `knitr`

As I said above, you can integrate everything you know about LaTeX into `knitr`—mathematical expressions, text formatting, whatever.

Compile the following set of text in your current .Rnw document just to check that LaTeX and `knitr` play nicely.
```asis
\documentclass[english, 11pt]{article}
\usepackage[utf8]{inputenc}

\begin{document}

\noindent Clever text.

\section{Some math}
In this section, we will write some Greek letters with impressive subscripts.
For instance, check out $\alpha_{i,j,k,\ell}$ and $\beta_{p,q,r,s,t,u,v,w}$.
Or, perhaps you prefer
\[
  \gamma_{12345} + \pi_0
\]

\end{document}
```

Okay, so you see that you can write whatever you like in LaTeX and `knitr` will comply/compile. Let's see what `knitr` actually brings.

## Chunks

One big change in `knitr`, relative to LaTeX, is the _chunk_ (their choice of words—not mine). You put R code into a chunk, and R will evaluate the code.

A chunk in `knitr` starts with the characters `<<>>=` on the top line and ends with an ampersat (`@`) on the last line, for instance,
```asis
<<>>=
# Let's add 2 and 2
2 + 2
@
```

Add the chunk above to your .Rnw file so that it now looks like
```asis
\documentclass[english, 11pt]{article}
\usepackage[utf8]{inputenc}

\begin{document}

\noindent Clever text.

<<>>=
# Let's add 2 and 2
2 + 2
@

\end{document}
```
and compile the PDF. You should now see the R code (and its comment) followed by the result of the R code (the number 4).

This simple example should give you a sense of what `knitr` can do: you can integrate your R code straight into your papers, assignments, diary entries, _etc._

## Chunk options

Inside of the `<<>>=`, you can specify options for the chunk. We will cover a few of the options now, and [this reference guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf) published by the ever-helpful folks at RStudio (pages 2–3) provides more information on the many options available for your chunks.^[A very strange sentence.]

The chunk options that I use most often (and coincidentally, the chunk options we will cover below) are:

- `engine`
- `eval`
- `include`
- `echo`

### The `engine` chunk option

I'm only covering this option because you will often see people write chunks that start with `<<R>>=`. The `R` here denotes the _engine_ (language), since you can use `knitr` for many programming/scripting languages. We will stick with `R` in our class. The following three chunks are equivalent because `knitr` defaults to the R engine.
```{R eval = F}
<<>>=
2 + 2
@
```
```{R eval = F}
<<R>>=
2 + 2
@
```
```{R eval = F}
<<engine = R>>=
2 + 2
@
```

### A quick aside

I going to start a new chunk of code that we will test with various chunk options. Here's the chunk
```{r, eval = F}
<<>>=
# Add two and two
2 + 2
# Set the seed
set.seed(12345)
# Generate two vectors of 100 N(0,1) random numbers
x <- rnorm(100)
y <- rnorm(100)
# Create a plot of x and y
plot(x, y)
@
```
which should produce the following output if you paste it into your .Rnw file and compile a PDF.
```{r, eval = T, echo = T, collapse = T}
# Add two and two
2 + 2
# Set the seed
set.seed(12345)
# Generate two vectors of 100 N(0,1) random numbers
x <- rnorm(100)
y <- rnorm(100)
# Create a plot of x and y
plot(x, y)
```

### The `eval` chunk option

Now let's check out the `eval` chunk option. Like many chunk options, you either set `eval` to `TRUE` or `FALSE` (for fans of minimalism: you can get away with `T` or `F`).

- `eval = TRUE` (the default) means R will evaluate the code in the chunk.
- `eval = FALSE` means R will not run the code in the chunk (it will still display the code).

The `eval = FALSE` option is a nice way to dispaly a lot of code without having to run it.

Since `eval = TRUE` is the default, we have already seen what it does. Now let's try `eval = FALSE` as an chunk option. Change the chunk's option to `<<eval = FALSE>>=` so that the chunk now looks like
```{r, eval = F}
<<eval = T>>=
# Add two and two
2 + 2
# Set the seed
set.seed(12345)
# Generate two vectors of 100 N(0,1) random numbers
x <- rnorm(100)
y <- rnorm(100)
# Create a plot of x and y
plot(x, y)
@
```

### The `include` chunk option

The `include` chunk option determines whether or not the chunk shows up in the final document.

- `include = TRUE` (the default) means the chunk will show up in the final document (PDF in our case)
- `include = FALSE` means the chunk will __not__ show up in the final document

Either way, R will evaluate the code in the chunk. The `include` option only determines what is _displayed_—not what is run (`eval` does that part).

Change the chunk option to `<<include = FALSE>>=`. The chunk's code and output should disappear from your screen.

`include` is helpful when you have a lot of intermediate steps that the reader does not need/want to see.

### The `echo` chunk option

The `echo` chunk option tells `knitr` whether you would like the code displayed above the output.

- `echo = TRUE` (the default) means the chunk's code will appear above its output.
- `echo = FALSE` means the chunk's code will not appear above its output, _i.e._ you will only see the output.

Try changing the code options in the chunk to `<<echo = FALSE>>=`. You should now only see the output without any preceding code, _i.e._,
```{r, echo = FALSE}
# Add two and two
2 + 2
# Set the seed
set.seed(12345)
# Generate two vectors of 100 N(0,1) random numbers
x <- rnorm(100)
y <- rnorm(100)
# Create a plot of x and y
plot(x, y)
```

The `echo = FALSE` option is helpful for writing papers/reports, since most people generally only care about results and do not want to see huge blocks of code between graphics.

## Combining options

To combine multiple options, separate the options with a comma, _e.g._,
```{r, eval = F}
<<R, eval = TRUE, echo = FALSE>>=
```
This use is a unnecessary, since two of the options are defaults, but you get the point.

## Naming your chunks

You can also name your chunks. In addition to strengthening your relationship with your chunks, naming them also helps you reference the figures and assists with troubleshooting. Place the chunk's name before you call any other chunk options.

You can also use the `fig.cap` option to place a caption below your figure. You need a non-empty argument to `fig.cap` in order for figure referencing to work.

Example:
```{r, eval = F}
<<myLittleChunk, fig.cap = "Plotting x and y", eval = TRUE, echo = FALSE>>=
# Set the seed
set.seed(12345)
# Generate two vectors of 100 N(0,1) random numbers
x <- rnorm(100)
y <- rnorm(100)
# Create a plot of x and y
plot(x,y)
@
Notice that Figure~\ref{fig:myLittleChunk} displays the relationship between $x$ and $y$.
```

I named my chunk `myLittleChunk` and gave the figure a caption `Plotting x and y`. After the chunk, I then referenced the figure using LaTeX's referencing function `\ref{}` and the chunk's name.^[LaTeX also offers a function called `\pageref{}` that gives you the page number, rather than the figure number.]

## Inline referencing

`knitr` has a pretty cool function called `\Sexpr{}`. So far everything we've done with `knitr` has been in a code chunk. `\Sexpr{}` lives outside of code chunks and allows you to access results/data in your text. This feature is useful (1) because code chunks break up the continuity of your document and (2) because you might want a simple number (_e.g._ the of households in your sample) that does not need a huge code block or a new paragraph.

We already have two objects called `x` and `y` lurking in our computers' memories. Did you know the means of $x$ and $y$ are appoximately `r round(mean(x), 3)` and `r round(mean(y), 3)`, respectively? Do you know how I know the means? `\Sexpr{}` told me.^[100% transparency: I'm using Rmarkdown, which actually uses a different literal expression that `\Sexpr{}`. Same idea.]

To use `\Sexpr{}`, you simply include `\Sexpr{}` in your text and write your R code inside of `\Sexpr{}`'s curly brackets. R and `knitr` do the rest.

For instance, to talk about the means of `x` and `y`, add the following line of code to your .Rnw file (somewhere below the chunk where we generate `x` and `y`) and recompile the PDF.
```{R, eval = F}
Throughout history, many people have wondered about the means of $x$ and $y$.
Today we can provide some answers. The mean of $x$ is approximately
\Sexpr{round(mean(x), 3)}. The mean of $y$ is approximately \Sexpr{round(mean(y), 3)}.
```

# References

You have several options for managing your references and making bibliographies in LaTeX (_e.g._, biber,^[Not Justin.] bibtex and biblatex). ShareLaTeX has a [very helpful discussion](https://www.sharelatex.com/learn/Bibliography_management_in_LaTeX) of your options and how to use them.
