---
title: "Section 1: Getting started with R"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Summary

In this section...

# Packages in R

Open RStudio.^[If you do not have the most recent versions of R (3.3.2, a.k.a. _Sincere Pumpkin Patch_) and RStudio (1.0.136), then please check out the directions in [Section 0](section00.html).]

While the base R installation helpful/powerful, R's true potential comes from combining combining the core installation with its many packages generated through collaboration (see [CRAN's list of packages](https://cran.r-project.org/web/packages/available_packages_by_name.html)^[CRAN stands for [the] Comprehensive R Archive Network]).

## Installing packages

Now, let's install a few packages that will prove useful this semester...

```{r, eval = F}
# Install the package named "dplyr"
install.packages("dplyr")
# Install the packages named "haven" and "readr"
install.packages(c("haven", "readr"))
```

A few things to notice here:

1. The name of the function `install.packages()` is plural regardless of the number of packages.
2. Each package's name is surrounded by quotes (_e.g._ `"haven"`). These quotation marks are how R knows the difference between characters/strings and objects (object hold values). R generally does not care whether you use single quote (`''`) or double quotes (`""`), but you should be consistent.
3. We can create a vector of packages using the combine function `c()`. Example: `c(1, 2, 3)` is a three-element vector whose elements are `1`, `2`, and `3`. Similarly, `c("haven", "readr")` is a two-element vector whose elements are the `"haven"` and `"readr"`. Vectors are a big deal in R.
4. The hashtag (`#`) is the symbol that creates comments in R.

## Loading packages

To check that the installations were successful, we will load the packages that we installed above.

R uses the `library()` function to load a package (we give the name of the package as the argument to the function).

```{r, eval = F}
library(dplyr)
library(haven)
library(readr)
```

Notice that we did not need to call the packages with quotations around their names (though it would still work).

# Loading data

Loading a dataset in R requires three things:

1. The path of the data file (where the data exist on your computer)
2. The name of the data file
3. The proper function for the type of dataset (_e.g._ different function for .csv and .dta files)

## File paths

R wants file paths as character vectors (_i.e._ the file path surrounded by quotations). For example, `"/Users/edwardarubin/Dropbox/Teaching/ARE212"` is the path to my folder for this course (on my computer).^[Windows users beware: when you copy the path from File Explorer, the slashes between folders may be in the wrong direction for R: you will need to either change the direction (from `\` to `/`) or double them (from `/` to `//`).]

To change the directory in R, use the `setwd()`. For instance, to change R's directory to my course folder

```{r}
setwd("/Users/edwardarubin/Dropbox/Teaching/ARE212")
```

To find R's current working directory, simply type `getwd()`:

```{r}
getwd()
```

There are a few ways to deal with file paths and directories. One common way is to use `setwd()` at the top of your script (or invoke the function whenever you need to change directories).

I find clearer to store the paths that I will use—at the start of my R script, I define the paths relevant to the files I use within the script. I prefer this method because it allows me to quickly update paths and easily access subfolders.

```{r}
# The path to my ARE 212 folder (ARE212)
dir_class <- "/Users/edwardarubin/Dropbox/Teaching/ARE212/"
# The path to my section 1 folder (Section01), which is inside my ARE 212 folder
dir_section1 <- paste0(dir_class, "Section01/")
```

## `dir()`

The function `dir()` allows you to see contents of a folder. `dir()` can help when you forget the name of the file you want. To see the contents of a folder, give the folder's file path to `dir()`:

```{r}
# Look inside my ARE212 folder (dir_class stores the path)
dir(dir_class)
# Look inside my section 1 folder (dir_section1 stores the path)
dir(dir_section1)
```

## Functions to load files

Blah blah blah
